Structuring Theories with Implicit Morphisms

### 2018-10-19 implicit.tex
finished for submission

The motivation behind implicit morphisms is to generalize the benefits of include declarations even further.
Sometimes an atomic morphism is conceptually similar to an inclusion: For example, we might want to declare $m:\cn{Division}\to\cn{Group}=\{\cn{divide} := \lambda x,y.x\circ y^{-1}, \;\ldots\}$ to show how division is ``included'' into the theory of groups.
Other times we have to use structures because we need to duplicate declarations but we would still like to use them like includes.
For example, we have to declare the theory of rings using two structures (for the commutative group and the monoid).
But we still want a ring to be implicitly converted into, e.g., a commutative group.

Our key idea is to use a commutative{\footnotemark} subdiagram of the \mmt diagram, which we call the \emph{implicit-diagram}.
It contains all theories but only some of the morphisms -- the ones designated as \emph{implicit}.
Because the implicit-diagram commutes, there can be at most one implicit morphism from $S$ to $T$ -- if this morphism is $M$, we write $\ipc{M}{S}{T}$.
\footnotetext{A diagram commutes if for any two morphisms $M,M'\in\Mo{S}{T}$, we have $M\equiv M'$.}
%Practically, this hasis a multi-graph whose nodes are the theories and whose edges are the atomic morphisms.
%We can extend the correspondence by identifying all paths with the composition of all edges along the path.
%Then concatenation of paths is just composition of morphisms, and the empty path from $T$ to itself is the identity morphism $\id{T}$.
%(This is well-defined because composition is associative and identity is neutral.)
%A diagram is commutative if any two paths between the same nodes are equal (with respect to $\equiv$), i.e., there is at most one morphism between any two nodes.

The implicit-diagram generalizes the inclusion relation $\harr$.
In particular, all identity and inclusion morphisms are implicit, and we recover $S\harr T$ as the special case $\ipc{\id{S}}{S}{T}$.
Moreover, the relation ``exists $M$ such that $\ipc{M}{S}{T}$'' is also an order.

Consequently, many of the advantages of inclusions carry over to implicit morphisms.
\begin{compactitem}
\item It is very easy to maintain the implicit-diagram, e.g., as a partial map that assigns to a pair of theories the implicit morphism between them (if any).
\item We can generalize the visibility of identifiers: If $\ipc{M}{S}{T}$, we can use all $S$-identifiers in $T$ as if $S$ were included into $T$.
Any $c\in\dom{S}$ is treated as a valid $T$-identifiers with definiens $M(c)$.
\item We can still use canonical identifiers $s?n$. Because there can be at most one implicit morphism $\ipc{M}{s}{T}$, using $s?n$ as an identifier in $T$ is unambiguous.
Crucially, we can use $s?n$ without bothering what $M$ is.
\end{compactitem}

@@

It remains to define which morphisms are implicit.
For that purpose, we allow \mmt declarations to carry \textit{attributes}.
Precisely, we add the following productions
\begin{grammar}
MDec   & Att\; MDec  & \text{attributed morphism} \\
Dec    & Att\; Dec   & \text{attributed declaration}\\
Att    & \keyword{implicit} \alt \ldots & \text{attributes}
\end{grammar}

The set of attributes is itself extensible, and the above grammar only list the one that we use to get started.
Additional attributes can be added when adding modularity principles.

@@

We only have to make two minor changes to the semantics to accommodate implicit morphisms.
The first change govern how we obtain implicit morphisms, the second one how we use them.

\paragraph{Obtaining Implicit Morphisms}
Based on the above attributes, we define the set $\IMo{S}{T}\sq\Mo{S}{T}$ of \textbf{implicit} morphisms to contain the following elements:
\begin{itemize}
 \item all declared morphisms $m:S\to T=\{\sigma\}$ whose declaration carries the attribute $\keyword{implicit}$,
\begin{nomodexp}
 \item all identity morphisms $\id{T}$,
 \item all compositions $M;N$ of implicit morphisms $M$ and $N$,
\end{nomodexp}
 \item all morphisms that additional language features designate as implicit based on the use of additional attributes.
\end{itemize}

To define well-formedness for our extended syntax, we say that adding an attribute to a declaration is only well-formed if there is (up to equality of morphisms) at most one implicit morphism for any pair of theories, i.e., $\IMo{S}{T}$ has cardinality $0$ or $1$.
Thus, the implicit morphisms form a thin broad subcategory of theories and morphisms --
in other words, a commutative diagram.

@@

\begin{example}[Includes (continued from Ex.~\ref{sem:incl})]\label{impl:incl}
For include morphism, we add the following definition: if theory $t$ contains the declaration $\icl{S}$, then the induced morphism from $S$ to $t$ is implicit.

Combined with composition morphisms, we see that all transitive includes between theories are implicit.
That corresponds to the intuition that anything that is include should be available directly.
\end{example}

\begin{example}
	For our morphism $\cn{DG2G}$ from Ex.~\ref{ex:dg2gimplicit}, this means that all symbols declared in the theory $\cn{DivisionGroup}$ (see Ex.~\ref{ex:thgroup}) are now visible in the theory $\cn{Group}$ with the definitions provided by $\cn{DG2G}$.
\end{example}

@@

\paragraph{Using Implicit Morphisms}
The intuition behind implicit morphisms is that all $S$-constants $c:E$ that can be mapped into the current theory via an implicit morphism $M:S\to T$ are directly available in $T$.
We can practically realize this by adding new defined constants $c:M(E)=M(c)$ to $T$.
However, physically adding definitions can be inefficient.
It is more elegant to modify the typing rules such that $\vdash_T c:M(c)=M(E)$.

To do that, we only have to make a small modification to the original rules of \mmt as presented in Sect.~\ref{sec:mmt}.
To illustrate how simple the modification is, we repeat the original rule first for comparison:
\[\rul{c:E[=e] \minn \flt{T}}{\vdash_T c[=e]:E}\]
Now our modified rule is
\[\rul{c:E[=e] \minn \flt{S}\tb \ipc{M}{S}{T}}{\vdash_T c=M(c):M(E)}\]

Note that the modified rule gives every constant a definiens.
This is a technical trick to subsume the original rule: if $c$ is already declared in $T$, we use $M=\id{T}$ and obtain $\vdash_T c=c:E$.
We write $\ov{E}$ for the expression that arises from $E$ by recursively replacing every $c$ with its definiens.

The following theorem is our central theoretical result.
It shows that the above modification has the intended properties:
\begin{theorem}[Conservativity of Implicit Morphisms]
Consider any combination of modular features from the examples above that includes at least implicit identity morphisms.

Then for well-formed diagrams and $S,T\in\Thy$ and $M\in\Mo{S}{T}$
\begin{compactenum}
 \item Whenever the original system proves $\vdash_T e=e':E$, so does the modified one.
 \item Whenever the modified system proves $\vdash_T e=e':E$, then the original inference system proves $\vdash_T \ov{e}=\ov{e'}:\ov{E}$.
\end{compactenum}
\end{theorem}

\begin{proof}
For the first claim, we proceed by induction on derivations.
We only need to consider case where the original rule was applied.
So assume it yields $\vdash_T c[=e]:E$, i.e., $(c:E[=e]) \minn \flt{T}$.
We apply the modified rule for the special case $\ipc{\id{T}}{T}{T}$.
The conclusion reduces to
\begin{compactitem}
 \item if $e$ is absent: $\vdash_T c=c:E$, which is equivalent to $\vdash_T c:E$,
 \item if $e$ is present: $\vdash_T c=e:E$ because $M(c)=M(e)$ according to the definition of $M(-)$.
\end{compactitem}

For the second claim, we proceed by induction on derivations.
We only need to consider the cases where the modified rule was applied.
So assume it yields $\vdash_T c=M(c):M(E)$ for $(c:E[=e]) \minn \flt{S}$ and $\ipc{M}{S}{T}$.
We distinguish two cases:
\begin{compactitem}
 \item $M(c)=c$, i.e., $c$ does not have a definiens:
   According to the definition of $M(-)$ this is only possible if $e$ is absent.
   According to the definition of $\flt{T}$, this is only possible if $c=S?n$ for $S\harr T$ (including the special case $S=T$).
   In that case, $\flt{S}\sq\flt{T}$ and $M$ is the include/identity morphism that maps all $S$-constants to themselves.
   Now applying the induction hypothesis to the well-formedness derivation of $E$ yields $\vdash_T c:\ov{E}$ as needed.
 \item $M(c)\neq c$, i.e., $c$ has a non-trivial definiens:
  Definition expansion eliminates $c$ in favor of $e$.
  Clearly $\ov{c}=\ov{M(c)}$, and we only have to show that $\vdash_T \ov{M(c)}:\ov{M(E)}$ That follows from the judgment preservation of morphisms.
\end{compactitem}
\end{proof}

>>>>

Our key idea is to use a commutative subdiagram of the \mmt diagram, which we call the \emph{implicit-diagram}.
It contains all theories but only some of the morphisms --- the ones designated as \emph{implicit}.
Because the implicit-diagram commutes, there can be at most one implicit morphism from $S$ to $T$ --- if this morphism is $i$, we write $\ipc{i}{S}{T}$.
%Practically, this hasis a multi-graph whose nodes are the theories and whose edges are the atomic morphisms.
%We can extend the correspondence by identifying all paths with the composition of all edges along the path.
%Then concatenation of paths is just composition of morphisms, and the empty path from $T$ to itself is the identity morphism $\id{T}$.
%(This is well-defined because composition is associative and identity is neutral.)
%A diagram is commutative if any two paths between the same nodes are equal (with respect to $\equiv$), i.e., there is at most one morphism between any two nodes.
The implicit-diagram generalizes the inclusion relation:
All identity and inclusion morphisms are implicit, and we recover the inclusion relation $S\harr T$ as the special case $\ipc{\id{S}}{S}{T}$.
And just like inclusion, the relation ``exists $i$ such that $\ipc{i}{S}{T}$'' is a preorder.

Consequently, many of the advantages of inclusions carry over to implicit morphisms:
\begin{compactitem}
\item It is very easy to maintain the implicit-diagram, e.g., as a partial map that assigns to a pair of theories the implicit morphism between them (if any).
\item We can generalize the visibility of identifiers: If $\ipc{i}{S}{T}$, we can use all $S$-identifiers in $T$ as if $S$ were included into $T$.
Any $c\in\dom{S}$ is treated as a valid $T$-identifiers with definiens $M(c)$.
\item We can use canonical identifiers $S?n$ without worrying about ambiguity. Because there can be at most one implicit morphism $\ipc{i}{S}{T}$, using $S?n$ as an identifier in $T$ is unambiguous.
%Crucially, we can use $S?n$ without bothering what $i$ is.
\end{compactitem}

@@

It remains to define which morphisms are implicit.
For that purpose, we allow \mmt declarations to carry \textit{attributes}:

\begin{definition}[Attributes for Implicitness]
We add the following productions
\begin{grammar}
MDec   & Att\; MDec  & \text{attributed morphism} \\
Dec    & Att\; Dec   & \text{attributed declaration}\\
Att    & \keyword{implicit} \alt \ldots & \text{attributes}
\end{grammar}
\end{definition}

The set of attributes is itself extensible, and the above grammar only lists the one that we use to get started.
Additional attributes can be added when adding modularity principles.

@@

We only have to make two minor changes to the semantics to accommodate implicit morphisms.
The first change governs how we obtain implicit morphisms, the second one how we use them.

\begin{definition}[Obtaining Implicit Morphisms]
We define the set $\IMo{S}{T}\sq\Mo{S}{T}$ of \textbf{implicit} morphisms to contain the following elements:
\begin{itemize}
 \item all declared morphisms $m:S\to T=\{\sigma\}$ whose declaration carries the attribute $\keyword{implicit}$,
\begin{nomodexp}
 \item all identity morphisms $\id{T}$,
 \item all compositions $M;N$ of implicit morphisms $M$ and $N$,
\end{nomodexp}
 \item all morphisms that additional language features designate as implicit based on the use of additional attributes.
\end{itemize}

Adding an attribute to a declaration is only well-formed if there is (up to equality of morphisms) at most one implicit morphism for any pair of theories.
\end{definition}

@@

\begin{example}[Includes (continued from Ex.~\ref{sem:incl})]\label{impl:incl}
For include morphisms, we add the following definition: if theory $T$ contains the declaration $\icl{S}$, then the induced morphism from $S$ to $T$ is implicit.

Combined with composition morphisms, we see that all transitive includes between theories are implicit.
That corresponds to the intuition that anything that is included is available by its original name.
\end{example}

\begin{example}
	For our morphism $\cn{DG2G}$ from Ex.~\ref{ex:dg2gimplicit}, this means that all symbols declared in the theory $\cn{DivGroup}$ (see Ex.~\ref{ex:thgroup}) are now visible in the theory $\cn{Group}$ with the definitions provided by $\cn{DG2G}$.
	For example, we can write $a/b$ in $\cn{Group}$, where $/$ refers to the identifier $\cn{DivGroup}?\cn{div}$.
\end{example}

@@

The intuition behind implicit morphisms is that all $S$-constants $c:E$ that can be mapped into the current theory via an implicit morphism $M:S\to T$ are directly available in $T$.
We can practically realize this by adding new defined constants $c:M(E)=M(c)$ to $T$.
However, physically adding definitions can be inefficient.
It is more elegant to modify the typing rules such that $\vdash_T c:M(c)=M(E)$ holds without any changes to $T$.

To do that, we only have to make a small modification to the original rules of \mmt as presented in Sect.~\ref{sec:mmt}.
To illustrate how simple the modification is, the following definition repeat the original rule first for comparison:

\begin{definition}[Using Implicit Morphisms]\label{def:useimpl}
We replace the rule
\[\rul{c:E[=e] \minn \flt{T}}{\vdash_T c[=e]:E} \tb \mwith\tb \rul{c:E[=e] \minn \flt{S}\tb \ipc{M}{S}{T}}{\vdash_T c=M(c):M(E)}\]
\end{definition}

Note that the modified rule gives every constant a definiens.
This is a technical trick to subsume the original rule: if $c$ is already declared in $T$, we use $M=\id{T}$ and obtain $\vdash_T c=c:E$.

The following theorem is our central theoretical result.
It shows that the modification made in Def.~\ref{def:useimpl} has the intended properties:
\begin{theorem}[Conservativity of Implicit Morphisms]
For well-formed diagrams and $S,T\in\Thy$ and $M\in\Mo{S}{T}$:
\begin{compactenum}
 \item Whenever the original system proves $\vdash_T e=e':E$, so does the modified one.
 \item Whenever the modified system proves $\vdash_T e=e':E$, then the original system proves $\vdash_T \ov{e}=\ov{e'}:\ov{E}$.
\end{compactenum}
Here $\ov{E}$ is the expression that arises from $E$ by recursively replacing every constant with its definiens.
\end{theorem}

\begin{proof}
For the first claim, we proceed by induction on derivations.
We only need to consider the case where the original rule was applied.
So assume it yields $\vdash_T c[=e]:E$, i.e., $(c:E[=e]) \minn \flt{T}$.
We apply the modified rule for the special case $\ipc{\id{T}}{T}{T}$.
The conclusion reduces to
\begin{compactitem}
 \item if $e$ is absent: $\vdash_T c=c:E$, which is equivalent to $\vdash_T c:E$,
 \item if $e$ is present: $\vdash_T c=e:E$ because $M(c)=M(e)$ according to the definition of $M(-)$.
\end{compactitem}

For the second claim, we proceed by induction on derivations.
We only need to consider the case where the modified rule was applied.
So assume it yields $\vdash_T c=M(c):M(E)$ for $(c:E[=e]) \minn \flt{S}$ and $\ipc{M}{S}{T}$.
We distinguish two cases:
\begin{compactitem}
 \item $M(c)=c$, i.e., $e$ is absent, and $\ov{c}=c$.
   According to the definition of $\flt{S}$, this is only possible if $S\harr T$ (including the special case $S=T$).
   In that case, $\flt{S}\sq\flt{T}$ and $M$ is the include/identity morphism that maps all $S$-constants to themselves.
   Now applying the induction hypothesis to the well-formedness derivation of $E$ yields $\vdash_T c:\ov{E}$ as needed.
 \item $M(c)\neq c$, i.e., $e$ is present or $M$ assigns a non-trivial value to $c$.
  Definition expansion eliminates $c$ in favor of $M(c)$, and thus $\ov{c}=\ov{M(c)}$.
  We only have to show that $\vdash_T \ov{M(c)}:\ov{M(E)}$ That follows from the judgment preservation of morphisms.
\end{compactitem}
\end{proof}

### 2018-10-19 intro.tex
finished for submission

\paragraph{Motivation}
Theory morphisms have proved an essential tool for managing collections of theories in logics and related formal systems.
They can be used to structure theories and build large theories modularly from small components or to relate different theories to each other \cite{asl,devgraphs,littletheories}.
Areas in which tools based on theories and theory morphisms have been developed include specification \cite{obj,hets}, rewriting \cite{maude}, theorem proving \cite{imps}, and knowledge representation \cite{RK:mmt:10}.

@@

These systems usually use a logic $L$ for the fine-granular formalization of domain knowledge, and a diagram $D$ in the category of $L$-theories and $L$-morphisms for the high-level structure of large bodies of knowledge.
This diagram is generated by all theories and morphisms occurring in a user's development.

@@

For example, a user might reference an existing theory $\cn{Monoid}$, define a new theory $\cn{Group}$ that extends $\cn{Monoid}$, define a theory $\cn{DivGroup}$ (providing an alternative formulation of groups based on the division operation), and then define two theory morphisms $\cn{G2DG}:\cn{Group}\darr \cn{DivGroup}:\cn{DG2G}$ that witness an isomorphism between these theories.
This would result in the diagram on the right.%
\footnote{Note that we use the syntactic direction for the arrows, e.g., an arrow $m:S\to T$ states that any $S$-expression $E$ (e.g., a sort, term, formula, or proof) can be translated to an $T$-expression $m(E)$. Models are translated in the opposite direction.}
%Crucially, $m(-)$ preserves typing and provability.

@@

\paragraph{Contribution}
At the highest level, our contribution is the observation that implicit morphisms form a sweet spot of a very simple language feature that has substantial practical uses.
We recommend using implicit morphisms in all theory morphism--based formalisms.

More concretely, we present a formal system for developing structured theories with implicit morphisms.
Our starting point is the \mmt language \cite{RK:mmt:10}, which already provides a very general setting for defining and working with theories and morphisms.
\mmt is foundation-independent in the sense that it allows embedding a large variety of declarative languages (logics, type-theories, etc.).
Therefore, all our results are foundation-independent --- users of a particular language $L$ can use \mmt directly as a structuring formalism for $L$ or can easily transfer our results to a dedicated implementation of $L$.

@@

\paragraph{Overview}
In Sect.~\ref{sec:mmt}, we present the syntax and semantics of \mmt.
Even though the \mmt language is not new, our presentation is an entirely novel contribution in itself: it is much simpler and more elegant than the original one in \cite{RK:mmt:10}.
%Based on this extensible definition, we can introduce several additional structuring mechanisms.
%Many of these have already been implemented in the \mmt tool but were not formally defined --- a formal definition of their semantics as a part of the \mmt language had previously proved too complicated to spell out elegantly.
%With our reformulation, this is now not only possible but very easy.
%As an example, we define include declarations, the most important special case of implicit morphisms.
%\medskip
%
%Secondly, we describe implicit morphisms as a novel feature of \mmt.
Crucially, this increase in simplicity allows spelling out the syntax and semantics of implicit morphisms, which we do in Sect.~\ref{sec:impl}, within a few pages.
In Sect.~\ref{sec:appl}, we present applications.
Finally we review related and future work in Sect.~\ref{sec:conc}.

>>>>

\paragraph{Motivation}
Theory morphisms have proved an essential tool for managing collections of theories in logics and related formal systems.
They can be used to structure theories and build large theories modularly from small components or to relate different theories to each other \cite{asl,devgraphs,littletheories}.
Areas in which tools based on theories and theory morphisms have been developed include specification \cite{obj,hets}, rewriting \cite{maude}, theorem proving \cite{imps,isabelle_locales}, and knowledge representation \cite{RK:mmt:10}.
Closely related concepts are used in both object-oriented (\emph{classes}) and (\emph{type classes}) functional programming languages.

@@

These systems usually use a logic $L$ for the low-level formalization of domain knowledge, and a diagram $D$ in the category of $L$-theories and $L$-morphisms for the high-level structure of large bodies of knowledge.

@@

For example, a document might reference an existing theory $\cn{Monoid}$, define a new theory $\cn{Group}$ that extends $\cn{Monoid}$, define a theory $\cn{DivGroup}$ (providing an alternative formulation of groups based on the division operation), and then define two theory morphisms $\cn{G2DG}:\cn{Group}\darr \cn{DivGroup}:\cn{DG2G}$ that witness an isomorphism between these theories.
This would result in the diagram on the right.%
\footnote{Note that we use the syntactic direction for the arrows, e.g., an arrow $m:S\to T$ states that any $S$-expression $E$ (e.g., a sort, term, formula, or proof) can be translated to an $T$-expression $m(E)$. Models are translated in the opposite direction.}
%Crucially, $m(-)$ preserves typing and provability.

@@

\paragraph{Contribution}
At the highest level, our contribution is the observation that implicit morphisms form a sweet spot of a very simple language feature that has substantial practical uses.
We recommend using implicit morphisms in all theory morphism--based formalisms.
More concretely, we present a formal system for developing structured theories with implicit morphisms.
Our starting point is the \mmt language \cite{RK:mmt:10}, which already provides a very general setting for defining and working with theories and morphisms.
\mmt is logic-independent, i.e., allows embedding a large variety of declarative languages (logics, type-theories, etc.).
Therefore, all our results can be directly applied to any language $L$ represented in \mmt or easily transferred to dedicated implementations of $L$.

@@

\paragraph{Overview}
In Sect.~\ref{sec:mmt}, we present the syntax and semantics of \mmt.
Even though the \mmt language is not new, our presentation is an entirely novel contribution in itself: it is much simpler and more elegant than the original one in \cite{RK:mmt:10}.
%Based on this extensible definition, we can introduce several additional structuring mechanisms.
%Many of these have already been implemented in the \mmt tool but were not formally defined --- a formal definition of their semantics as a part of the \mmt language had previously proved too complicated to spell out elegantly.
%With our reformulation, this is now not only possible but very easy.
%As an example, we define include declarations, the most important special case of implicit morphisms.
%\medskip
%
%Secondly, we describe implicit morphisms as a novel feature of \mmt.
Crucially, this increase in simplicity allows spelling out the syntax and semantics of implicit morphisms, which we do in Sect.~\ref{sec:impl}, within a few pages.
In Sect.~\ref{sec:appl}, we present applications.
Finally we discuss related and future work in Sect.~\ref{sec:conc}.

### 2018-10-19 macros/listings.tex
typos
finished for submission

\usepackage{listings}\lstset{basicstyle=\sf\footnotesize,columns=fullflexible}

\lstnewenvironment{code}[1][mathescape=false]{\lstset{language=XML, #1}}{}

\newcommand{\delima}{\mid\negmedspace \mid}
\newcommand{\delimb}{\delima\negmedspace \delima}
\newcommand{\GS}{\ensuremath{\textcolor{RedViolet}{\delimb\negmedspace\delimb}}}
\newcommand{\RS}{\ensuremath{\textcolor{Green}{\delimb}}}
\newcommand{\US}{\ensuremath{\textcolor{Turquoise}{\delima}}}

\lstdefinestyle{mmt}{	
	keywordstyle=\color{Emerald},
	columns=flexible
}
\lstdefinelanguage{mmt}{
	morekeywords={theory,view,include,namespace,rule,role,implicit,import},
	morecomment=[l][\color{Gray}]{//},
	% morecomment={[s]{//}{❚},[s]{//}{❙},[s]{//}{❘}}
}
\lstnewenvironment{mmtcode}[1][language=mmt]{\lstset{language=mmt,style=mmt,inputencoding=utf8,escapechar=\&,#1}}{}
\lstnewenvironment{nmmtcode}[1][language=mmt]{\lstset{xleftmargin=1.2em,numbers=left,language=mmt,style=mmt,inputencoding=utf8,escapechar=\&,#1}}{}
\lstdefinestyle{scala}{	
	keywordstyle=\color{BurntOrange}
}
\lstdefinelanguage{scala}{
	emphstyle=\underbar,
	classoffset=0,keywordstyle=\bfseries\color{BurntOrange},
	morekeywords={case,class,object,trait,extends,match,if,else,def,implicit,import,val,var,this,override,with,return,abstract},
	classoffset=1,keywordstyle=\color{blue},
	morekeywords={OMA,OMS,OMBIND,OMV,OMID,Context,VarDecl,Term,LocalName,GlobalName,MPath,Solver,Stack,Rule,
	History,Inhabitable,Subtyping,Equality,Typing,Common,Judgment,FormationRule,CheckingRule,IntroductionRule,
	EliminationRule,InferenceRule,TypingRule,ComputationRule,TypeBasedEqualityRule,TermBasedEqualityRule,
	TermHeadBasedEqualityRule,Continue,CheckingCallback,Simplifiability,Simplify,RecurseOnly,SubtypingRule,
	VarianceRule,TypeBasedSolutionRule,DerivedDeclaration,StructuralFeature,Elaboration,FormatBasedExtension,
	Marker,ExtendedCheckingEnvironment,DeclaredModule,Universe,InhabitableRule,UniverseRule,ElementContainer,Declaration},
	classoffset=2,keywordstyle=\color{PineGreen},
	morekeywords={Option,Some,None,Boolean,String,List,Int,true,false,Nil},
	classoffset=0,
	morecomment=[l][\color{Gray}]{//},
	morecomment=[s][\color{Gray}]{/*}{*/},
	morecomment=[s][\color{OliveGreen}]{/**}{**/},
	morestring=[b][\color{Purple}]",
}
\lstnewenvironment{scalacode}[1][language=scala]{\lstset{language=scala,style=scala,inputencoding=utf8,escapechar=\&,#1}}{}
\lstnewenvironment{nscalacode}[1][language=scala]{\lstset{xleftmargin=1.2em,numbers=left,language=scala,style=scala,inputencoding=utf8,escapechar=\&,#1}}{}

%\newunicodechar{→}{\ensuremath{\rightarrow}}
\lstset{
	literate={→}{\ensuremath{\rightarrow\ }}1
			{⟶}{\ensuremath{\rightarrow\ }}1
			{Π}{\ensuremath{\Pi}}1
			{⟹}{\ensuremath{\Longrightarrow\ }}1
			{λ}{\ensuremath{\lambda}}1
			{⟨}{\ensuremath{\langle}}1
			{⟩}{\ensuremath{\rangle}}1
			{❚}{\GS}1
			{❙}{\RS}1
			{❘}{\US}1
			{Σ}{\ensuremath{\sum}}1
			{…}{\ldots}1
			{×}{\ensuremath{\times}}1
			{π}{\ensuremath{\pi}}1
			{⨁}{\ensuremath{\bigoplus}}1
			{↪}{\ensuremath{\hookrightarrow}}1
			{↩}{\ensuremath{\hookleftarrow}}1
			{⊕}{\ensuremath{\oplus}}1
			{∅}{\ensuremath{\emptyset}}1
			{★}{\ensuremath{\star}}1
			{ℕ}{\ensuremath{\mathbb N}}1
			{∘}{\ensuremath{\circ}}1
			{∀}{\ensuremath{\forall}}1
			{≐}{\ensuremath{\doteq}}1
			{⊦}{\ensuremath{\vdash}}1
			{⁻}{\ensuremath{^-}}1
			{¹}{\ensuremath{^1}}1
}

>>>>

\usepackage{listings}

\lstdefinelanguage{mmt}{
	morekeywords={theory,namespace,rule,role,import,implicit,definitional,renaming},
	morecomment=[l][\color{Gray}]{//},
	% morecomment={[s]{//}{❚},[s]{//}{❙},[s]{//}{❘}}
}
\lstnewenvironment{mmtcode}[1][language=mmt]{\lstset{language=mmt,keywordstyle=\bfseries,basicstyle=\ttfamily,mathescape,columns=fixed,#1}}{}
